Iterative Monomorphisation

### 2024-03-20 iterative_monomorphisation.tex
edits

The set of substitutions generated in this way is then iterated over, and each substitution is tentatively applied to all polymorphic type arguments. The application of a substitution to a tuple of polymorphic type arguments takes place only if least one type variable in the tuple is instantiated by the substitution. This condition filters out all substitution application that would have otherwise left all type arguments unchanged.

%[If we assume there are after \(k\) iterations \(N_k\) total type arguments split evenly between monomorphic and polymorphic type arguments, all in a single formula. There can be up to \((\frac{N_k}{2})^2\) successful matches yielding as many substitutions. Each substitution is then applied to each polymorphic type argument for a total of \((\frac{N_k}{2})^3\) new type arguments. If half of these new type arguments are monomorphic and the other half is polymorphic, then the next iteration will begin with \(N_{k+1} = (\frac{N_k}{2})^{3}\) evenly split type arguments. Meaning that the total number of type arguments on the \(k^\text{th}\) iteration can reach \(N_k = (\frac{N_0}{2})^{3^k}\). This paragraph is wrong on several points, it can be fixed but it would be annoying, the main idea is that it can be shown relatively easily in a semi-formal way that the total number of type arguments is doubly exponential in the number of iterations.]

The total number of type arguments can increase cubicly in the number of type arguments at each iteration and can therefore grow doubly exponentially in the number of iterations. Depending on the shape and size of the input problem and the number of iterations performed, this may be problematic. Introducing bounds can help solve this potential issue. The limit on the number of new monomorphic type arguments (in \textcolor{ourblueviolet}{blue}) has three components (similar bounds are used for the non-monomorphic type arguments):
\begin{enumerate}
   \item \textcolor{ourblueviolet}{\MonoCap} is an absolute limit on the total number of new type arguments.
   \item \textcolor{ourblueviolet}{\MonoMult} is used to allow the total number of type arguments to increase by a certain proportion of the current number of type arguments.
   \item \textcolor{ourblueviolet}{\MonoFloor} balances out \textcolor{ourblueviolet}{\MonoMult}, it guarantees that a minimum number of new type arguments even if the total number of type arguments is low.
\end{enumerate}

It is possible to either generalise or further refine these bounds. The maximum number of newly generated type arguments is fixed per formula, per iteration. This algorithm can easily be modified such that the limit on new type arguments is fixed per iteration. Conversely, new type arguments can be limited per symbol, per formula, per iteration.

>>>>

The set of substitutions generated in this way is then iterated over.
Each substitution is applied to all polymorphic type argument tuples such that at least one type variable in the tuple is instantiated by the substitution. %This condition filters out all substitution application that would have otherwise left all type arguments unchanged.

The total number of type arguments can increase cubicly in the number of type arguments at each iteration and can therefore grow doubly exponentially in the number of iterations. We sketch an example of such growth.
%
If we assume that there are after \(k\) iterations \(N_k\) total type arguments split evenly between monomorphic and non-monomorphic type arguments, all in a single formula, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions. Each substitution is then applied to each non-monomorphic type argument for a total of \(\bigl(\frac{N_k}{2}\bigr)^3\) possible new type argument tuples. If half of these new type arguments are monomorphic and the other half are non-monomorphic, then the next iteration will begin with \(N_{k+1} = \bigl(\frac{N_k}{2}\bigr)^{3}\) evenly split type argument tuples. Therefore, the total number of type arguments on the \(k\)th iteration can reach \(N_k = \bigl(\smash{\frac{N_0}{2}\bigr)^{3^k}} \cdot \bigl(\frac{1}{2}\bigl)^{3k}\).

Depending on the shape and size of the input problem and the number of iterations performed, the doubly exponential growth may be problematic. Introducing bounds addresses this potential issue. The limit on the number of new monomorphic type arguments (in \textcolor{ourblueviolet}{blue}) has three components:
\begin{enumerate}
   \item \textcolor{ourblueviolet}{\MonoCap} is an absolute limit on the total number of new type argument tuples.
   \item \textcolor{ourblueviolet}{\MonoMult} is used to allow the total number of type argument tuples to increase by a certain proportion of the current number of type argument tuples.
   \item \textcolor{ourblueviolet}{\MonoFloor} balances out \textcolor{ourblueviolet}{\MonoMult}. It prevents \textcolor{ourblueviolet}{\MonoMult} from inhibiting new type argument tuple generation if the total number of type argument tuples is low.
\end{enumerate}

Similar bounds are used for the non-monomorphic type arguments.
It is possible to either generalise or further refine these bounds. The maximum number of newly generated type arguments is fixed per formula, per iteration. This algorithm can easily be modified such that the limit on new type arguments is fixed per iteration. Conversely, new type arguments can be limited per symbol, per formula, per iteration.
